Cheminformatics, the Computer Science of Chemical Discovery, Turning Open-Source

### 2012-02-07 resubmit/cheminformatics_CACM_revision.tex

\documentclass{sig-alternate}

@@

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists that led as one of its major accomplishments to the
  sequencing of the human genome. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. While researchers in both academia and industry have made important contributions to this field since the 1960s (if not before), new and exciting research opportunities now exist because of an ``opening'' of data and software that began around the year 2000.  Since that time -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- researchers have gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that might have been prohibitive to tackle a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences produce information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

So why is chemical information important and why should we care about
its public availability -- and how does this relate to the field of
computer science? We note that though cheminformatics has a broad
applicability, ranging from agrochemical research to the design of
novel materials, we have chosen to use drug discovery as the context
of this article due to its central relevance for human well being.

Given that the art and science of drug discovery is based around small
molecules, techniques and algorithms to handle, analyse and interpret
chemical structure information are a core component of the methods of drug
discovery. Public availability of chemical information and
tools to handle it are important as drug discovery research is slow,
expensive and prone to failure. The more information available to each
researcher, facilitating the development of novel treatments, the
better the chances of avoiding the many causes of attrition. The
relevance of computer science to this research becomes clear when we
consider the quantity and types of data available -- currently a
single data base such as PubChem contains more than 34 million
chemical structure records, along with an even larger number of
\emph{annotations} (comments on the structure that are database
searchable), such as synonyms (different names for the same molecule),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and so on.  Simply considering the synonym
list, PubChem stores 50.1 million synonym for 19.6 million compounds.

The design of data structures to handle chemical structures and
associated information, as well as subsequent data mining are areas
where expertise in computer science is urgently needed. (For a recent,
detailed introduction to cheminformatics, written for computer
scientists, see \cite{brown2009}).

Cheminformatics comprises different areas, which broadly speaking can
be divided into three fields: \emph{capturing data} (using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications); \emph{storing data} (designing database
schemas, devising ontologies) and \emph{mining data} (such as for the
prediction of biological activities of compounds).  However, the type
of information being analyzed is different from biological
information: while bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. In the case of the
former, information can very frequently be represented as
one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter however, chemical
structures may possess rings, branches as well as multiple valid
representations of the same molecule (such as \emph{tautomers}, where
hydrogen atoms can be positioned in different places of the
molecules), potentially giving rise to ambiguities. Hence, chemical
structures are often reported to be more difficult to standardize and
algorithms dealing with them must take this into account.

To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The representations range
from a graph (Fig.~\ref{figure:chemical-structures}A) that contains
much implicit information (such as unlabeled nodes represent carbons
with valencies satisfied by hydrogens) to a fully labeled graph,
where all information is explicit
(Fig.~\ref{figure:chemical-structures}B). The former is what is
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations, to discover patterns in large-scale
datasets \cite{horst2009}; this information can also be determined by simulations such as molecular dynamics.

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left), a labeled graph (top
  right; 'A' corresponds to aromatic atoms, 'O' to oxygens and 'On' to
  oxygens or nitrogens. Unlabeled nodes have an implicit label of 'C'), a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

Although graphs implicitly encode the 3D structure of a molecule (when
combined with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure.

In addition, the 3D structure may have
a ``secondary'' geometrical arrangement of features (such as the
presence of a right-handed helix) which cannot be encoded in the
graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}C) that make the 3D arrangement
of atoms and bonds explicit. Fig.~\ref{figure:chemical-structures}D
goes further and represents a molecular surface, encoding some
property (such as lipophilicity, the ability to dissolve in non-polar
solvents) that varies over the surface.

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Fig.~\ref{figure:chemical-structures} shows a
hierarchy of representations; when searching for substructures, a 2D
representation suffices, but when exploring protein binding, a 3D
structure is vital. Figure \ref{fig:structure-classes} in the Appendix
highlights yet another subtlety where the number and types of atoms
and their connectivity may not suffice to uniquely define the
structure.

Given that a fundamental principle of the field is that \emph{similar
molecules exhibit similar properties} \cite{Johnson:1990qf}, the
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
even after one has selected a representation one faces the challenge
of balancing computational costs with the utility of the
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers would let us accurately
predict many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error - in that an objective function
(usually root mean square error or percentage correct) is defined and
descriptors and modeling approaches are iteratively adapted to
optimize the objective function . In other words, there is no
underlying theory to explain or suggest optimal approaches.

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous problems.
For the field of cheminformatics to flourish, we need closer
collaboration between chemists (or, more generally, life scientists)
and computer scientists -- where the former need to be able to pose
their problems in a way relevant for practical applications; and where
the latter are able to devise ways of capturing, storing, and
analyzing chemical data which achieve optimal balances between the
different risks in generalizable encodings and complexity (space and
time). The goals of cheminformatics are to support better chemical
decision making by 1. storing and integrating data in maintainable
ways, 2.  providing sufficient open standards and tools allowing
applications and data to be used across heterogeneous platforms, and
3. mining the many chemical property spaces in a time- and
space-efficient way. We will discuss all three goals in the following
sections.

@@

This main article will highlight current topics in
cheminformatics, presenting the computational and algorithmic problems
faced in the field and how modern CS research can contribute to their
solution.  We have also provided two appendices: a brief history of cheminformatics, and a detailed list of open problems.

We structure the main article around ``risk minimization'' of drug
discovery (a slow and expensive process)---how to minimize the chances
of a small molecule failing, due to poor physical, chemical or
biological properties, in the various research and development stages
as a drug candidate, and how to successfully employ cheminformatics
methods in these efforts. As an example, a molecule must be soluble
and show a certain degree of bioavailability to be considered as a
drug candidate. In cases where these properties are poor, a
cheminformatics approach can suggest replacement of certain functional
groups, that maintain potency but improve the solubility and
bioavailability.  Table~\ref{table:properties} (in the Appendix)
summarizes a number of properties a drug candidate must satisfy to be
considered therapeutically useful (such as being appropriately
soluble, exhibiting minimal toxicity and being selective for the
intended target) and sketches the role of cheminformatics at each
stage of drug development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
these areas poses its unique computer science challenges.

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Integration of these disparate
data is essential for researchers to obtain the fullest possible
perspective on what is presently known, tracking ongoing advances in
science as these become available. Still, integration of data
across chemical databases remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume.
Given the requirement to ``make sense'' of as much chemical data
as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change
from database to database. Hence, molecular graphs must be captured
in a machine-readable fashion.

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine-readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. Considering protonation states
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format as defined by Weininger 1988\cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the data points contained in a
particular database to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the structures in the database, while in similarity searching,
the database is searched for structures similar (in some structure or
property space) to a provided search, or query, structure. Chemical
search packages are often implemented and optimized for a given
database technology, for example, the OrChem package is an Open Source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In the former, each bit position corresponds to a
distinct substructure such as a \emph{functional group} (a connected
set of atoms that affects the characteristics of the chemical
reactions of the molecule). Examples include MACCS and PubChem
keys. The latter class are designed such that substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments, and so on. Examples include
ECFPs (Extended Connectivity Fingerprints, which use local topological
information) and Daylight fingerprints (which are ``folded'' to
optimize information density and screening speed).

Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query. This is
achieved by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, comparing binary strings can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, in which case a similarity metric such as the
Tanimoto coefficient is used to compare the query and target
fingerprints. While fast, heuristics have been developed
\cite{Swamidass:2007ve} to further speed up similarity searches.

@@

In the quest for the discovery of novel therapeutic agents, it is
essential to study molecules as they are situated in a biological
context, as metabolites, cofactors for enzymatic reactions, hormones
and signalling molecules. Thus we need to adequately represent and
organise chemical data across biological databases such as those
containing pathways, protein information or biological
samples. Integration and processing of data from such disparate
domains underlies ``systems'' level biological research.  This leads
to additional challenges that standard chemical structural
representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class
of compounds behave in a certain context in a biological system rather
than a single molecule with a fully-specified structure -- and for
similar reasons, chemical concepts such as \textit{groups} (parts of
molecules); secondly, the need to represent \textit{non-structural}
groupings of compounds of interest, since compounds may bind to the
same or similar targets or they may have similar biological functions
such as acting as an anti-neoplastic -- preventing the development of
tumors. With the many different biological databases needing to refer
to these different chemical entities in an organized and standardized
fashion to enable cross-database integration for a whole-systems
perspective, \textit{ontologies} and other semantic technologies are
being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within biological
systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases.
Ontologies are based on logical languages of varying levels of
expressivity, accompanied by sophisticated reasoning algorithms. They
can be used to construct various sorts of semantic similarity
(e.g. similarity in function, similarity in application) which
complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical structure
information with the wider logic-based ontology information, which
will allow combined automated reasoning and querying over the
different domains encoded in a growing number of interlinked
ontologies. % \cite{hastingsowled2010}.
Enabling automated inferencing in such heterogenous linked systems
brings us closer to achieving a systems level understanding of small
molecule activity \cite{Oprea:2007fk}.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules exhibit
similar properties), the goal of any modeling approach is to capture
and characterize correlations between structural features and
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using the
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually delineated by similarity
to the training set.

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. For
example, labeled graphs (Figure 1B), and their associated
characterizations are easily accessible to computer scientists, yet
such features miss significant physicochemical features (surface
properties such as charge, flexibility and so on). At the same time,
it can be difficult to objectively quantify many chemical aspects of a
molecule, such that the resultant descriptors are suitable for
predictive modeling.  Hence, the choice of a chemical descriptor
should by no means be treated as a ‘solved’ problem. For a more
detailed discussion the reader is referred to more comprehensive
textbooks \cite{todeschini2000,faulon2010}.

As noted above, molecular graphs can be transformed to numerical
vector representations ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.

Both methods have advantages and disadvantages. The vector approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem is
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and one must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels (such
as a multiple common substructure based kernel). These challenges can
make kernel based methods prohibitive on very large datasets.

>>>>

\documentclass{../sig-alternate}

@@

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists that led to the
  sequencing of the human genome and other accomplishments. However,
  few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  drug discovery to agrochemicals and materials design. 
  While researchers in both academia and industry have made important 
  contributions to this field for decades, new and 
  exciting collaborative opportunities have arisen from an ``opening'' of 
  data and software as an effect of changing mindsets,
  policy changes, 
  and chemists volunteering time
  for ``Open Science''.  Researchers have gained access to
  freely available open source software packages and open databases of tens of millions
  of chemicals, allowing academic chemists to confront a variety of 
  algorithmic problems whose solutions will be critical to address 
  current challenges ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
  In this paper, we give a broad overview of the field of cheminformatics with a
  focus on open questions and challenges. 
\end{abstract}

@@

Novel technologies in the life sciences produce information at an
ever increasing rate, with public data stores such as that of the
European Bioinformatics Institute (EBI) containing in the order of 10
petabytes of biological information.  Until recently, the same 
had not been true for
chemical information, but in 2004 a
large public small molecule structure repository (PubChem) became
freely available, and this was soon followed by other databases. 
Similarly, while many of
the foundational algorithms of cheminformatics have been described
since the 1950s, open source software implementing many of
these algorithms have only become accessible in the last ten to fifteen
years \cite{faulon2010}.

So why is chemical information important and why should we care about
its public availability -- and how does this relate to the field of
computer science? Cheminformatics is widely used in fields 
from agrochemical research to the design of
novel materials -- but we have chosen to use drug discovery as the context
of this article due to its central relevance for human well-being.
The art and science of drug discovery centers around small
molecules, with a core component of methods involving 
techniques and algorithms to handle, analyse and interpret
chemical structure information. 
Unfortunately, drug discovery research is slow, expensive and prone to failure. 
Public availability of chemical information and
tools to handle it are important, as the more information available to each
researcher, the better the chances of avoiding the many causes of attrition. 
Computer science research is highly relevant to managing the \textit{volume} 
and \textit{complexity} of chemical information. 
A single database such as PubChem contains more than 34 million
chemical structure records), along with an even larger number of
\emph{annotations}, such as synonyms (different names for the same molecule; 
PubChem stores 50.1 million synonym for 19.6 million compounds),
known targets (drugs with which the molecule is known to interact),
mode of action (how the molecule interacts with its targets),
regulatory approval status and much more.

While bioinformatics often deals with sequences, the
domain of cheminformatics is chemical structures. 
In the case of the former, information can frequently be represented as
one-dimensional strings, which are relatively easy to handle
computationally. In the case of the latter, however, chemical
structures are complex graphs that may possess rings and branches, 
and there may be multiple valid representations of the same molecule 
that need to be considered for particular algorithms 
(such as \emph{tautomers}, where hydrogen atoms are positioned at 
different places in the molecule). Hence, chemical
structures are more difficult to standardize and
algorithms dealing with them must take this into account.
To illustrate the difficulty of standardizing chemical structures,
different representations of the same chemical structure are shown in
Figure~\ref{figure:chemical-structures}. The top left graph contains
much implicit information, since unlabeled nodes represent carbons
with valencies satisfied by hydrogens, while the top right graph is 
fully labeled. The former is what is 
usually exchanged informally between chemists, but to actually mine
chemical structures computationally, something like the latter is
required.  For example, graph mining algorithms can be applied to
labeled graph representations to discover patterns in large-scale
datasets \cite{horst2009}; this information can also be determined 
by simulations such as molecular dynamics.

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (top left); a labeled graph (top
  right -- `A' corresponds to aromatic atoms, `O' to oxygens and `On' to
  oxygens or nitrogens, unlabeled nodes have an implicit label of `C'; a 3D
  conformation (bottom left) and a surface representation (bottom right). All show differing,
  valid aspects of the same molecule, and are suited for
  different purposes, with representation choice guided by
  intended use.}
\label{figure:chemical-structures}
\end{figure}

Although graphs implicitly encode the 3D structure of a molecule (when
combined with knowledge about favored bond angles and distances), many
different low energy 3D structures, or conformers, may be consistent
with the same graph structure. In addition, the 3D structure may have
a ``secondary'' geometrical arrangement of features (such as the
presence of a right-handed helix) which cannot be encoded in the
graph. Thus we have 3D representations
(Fig.~\ref{figure:chemical-structures}, bottom) that make the 3D arrangement
of atoms and bonds explicit. The representation at the bottom right 
goes further, showing a molecular surface and some
property that varies over the surface -- such as lipophilicity, the ability to dissolve in non-polar
solvents.
Even though some structure representations contain more explicit
information than others, they are all equally valid,  suited for
different problems. When searching for substructures, a 2D
representation suffices, but when exploring protein binding, a 3D
structure is vital. Figure \ref{fig:structure-classes} in the Appendix
highlights another subtlety where the number and types of atoms
and their connectivity may not suffice to uniquely define a
structure.

A fundamental principle of cheminformatics is that \emph{similar
molecules exhibit similar properties} \cite{Johnson:1990qf}. The
choice of representation is key in determining how such similarities
are evaluated, and thus the effectiveness of subsequent analyses. But
there is a further challenge in balancing computational costs with 
the utility of a
representation. For example, a full 3D description of a molecule,
taking into account all possible conformers, would allow accurate
prediction of many properties. But the size of this representation and the
time required to evaluate it would be prohibitive. Instead, can we
obtain accurate predictions from a subset of conformers? Or, can we
obtain comparable accuracy by using a 2D representation? And if so,
what type of labels are required? Currently, many of these questions
are answered using trial and error through definition of an objective function
(usually root mean square error or percentage correct) and
iterative adaptation of descriptors and modeling approaches to
optimize the objective function. There is no
unifying theory to explain or suggest optimal approaches in all cases.

The goals of cheminformatics are to support better chemical
decision making by (1) storing and integrating data in maintainable
ways, (2) providing sufficient open standards and tools allowing
applications and data to be used across heterogeneous platforms, and
(3) mining the many chemical property spaces in a time- and
space-efficient way. For the field of cheminformatics 
to flourish, we need closer
collaboration between chemists 
and computer scientists -- where the former need to be able to pose
their problems in a way relevant for practical applications, and where
the latter are able to devise ways of capturing, storing, and
analyzing chemical data which achieve an optimal balance between space,
performance and complexity. For a recent,
detailed introduction to cheminformatics, written for computer
scientists, see \cite{brown2009}.
Cheminformatics comprises different areas, which can broadly
be divided into three fields: \emph{capturing data} using lab
notebooks or potentially using formats such as the Chemical Markup
Language for publications; \emph{storing data} -- designing database
schemas, devising ontologies -- and \emph{mining data} such as for the
prediction of biological activities of compounds.  
We will discuss all of these aspects in the following
sections.

@@

This main article will highlight current topics in
cheminformatics, presenting computational and algorithmic problems
and indicating how modern CS research can contribute to their
solution.  We have also provided two appendices: a brief history of 
cheminformatics, and a list of open problems.

We use as running example cheminformatics methods for 
``risk minimization'' in drug
discovery -- minimizing the chances
of a small molecule failing, due to poor physical, chemical or
biological properties, during the various research 
and development stages
as a drug candidate. As an example, a molecule must be soluble
and show a certain degree of bioavailability to be considered as a
drug candidate. In cases where these properties are poor, a
cheminformatics approach can suggest replacement of certain functional
groups (connected
sets of atoms that affect the characteristics of the chemical
reactions of the molecule) 
to maintain potency but improve the solubility and
bioavailability.  Table~\ref{table:properties} (in the Appendix)
summarizes a number of properties a drug candidate must satisfy to be
considered therapeutically useful 
and sketches the role of cheminformatics at each
stage of drug development.
%
\subsection{Representing and Searching Structures}
\label{sec:databases}
%
Most cheminformatics applications rely on large databases of chemical
structures, their properties, and relationships to, \textit{e.g.},
biological targets.  Organizing and maintaining such databases, as
well as searching and clustering similar structures together, are
essential to enable many scientific applications. Yet, each one of
these areas poses computer science challenges.

Chemical and bioactivity
databases such as ChEMBL and PubChem are freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points. Integration of these disparate
data is essential for researchers to obtain the fullest possible
perspective on what is presently known, tracking ongoing advances in
science as these become available. Integration of data
across chemical databases is a challenge due to sheer data volume and
due to the difficulties in normalization of chemical and bioactivity
data. Molecular graphs must be captured
in a machine-readable fashion. It is also necessary to be able to 
search in these data, for example multi-labeled graphs, where graph labels can change
from database to database. Not only can labels differ between databases, 
but a chemical graph can be encoded in multiple ways, depending on how 
the nodes are ordered, resulting in multiple representations of the same
molecule. Considering protonation states 
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format as defined by Weininger in 1988 
\cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the search molecule 
to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the search structure, while in similarity searching,
structures are retrieved that are similar (in some structure or
property space) to the provided search structure. Chemical
search packages are often implemented and optimized for a given
database technology, for example, the OrChem package is an Open Source
chemical search package for the Oracle database \cite{rijnbeek2009}.

Graph substructure matching is a variant of \emph{graph isomorphism},
which is widely believed to be computationally
intractable~\cite{cordella2001}. To execute a graph isomorphism search
across a full chemical database of thousands or even millions of
structures is infeasible \cite{Weininger:2011ly}. Speedups can be
obtained via the use of heuristics such as structural
\emph{fingerprints}. Fingerprints encode characteristic features of a
given chemical structure, usually in a fixed-length
bitmap. Fingerprints fall broadly into two categories: structure keys
and hashed keys. In structure keys, each bit position corresponds to a
distinct substructure such as a functional group. Examples include MACCS and PubChem
keys. Hashed keys are those in which substructural patterns
are represented as strings and then hashed to a random bit
position. As a result, a given position can encode multiple
substructures. The advantage of such fingerprints is that they can
cover an arbitrarily large collection of substructures such as ``paths
of length $N$,'' or circular environments. Examples include
ECFPs (Extended Connectivity Fingerprints, which use local topological
information) and Daylight fingerprints (which are ``folded'' to
optimize information density and screening speed).
Given a binary fingerprint, we can first ``pre-screen'' a database, to
ignore molecules that cannot possibly match the query,
 by requiring that all bits in a query fingerprint must also
be present in the target fingerprint. Since the target fingerprints
are pre-computed, this check can be performed extremely
rapidly on modern hardware. As a result, we apply the actual
isomorphism test only on those molecules that pass the
screen. Fingerprints can also be used to rapidly search databases for
similar molecules, by using a similarity metric such as the
Tanimoto coefficient to compare the query and target
fingerprints. Additional heuristics have been developed
\cite{Swamidass:2007ve} to further speed up similarity searches.

@@

In the quest for the discovery of novel therapeutic agents, it is
essential to study molecules as they are situated in a biological
context, as metabolites, cofactors for enzymatic reactions, hormones
and signalling molecules. Thus we need to adequately represent and
organise chemical data across biological databases such as those
containing pathways, protein information or biological
samples. Integration and processing of data from such disparate
domains underlies ``systems'' level biological research.  This leads
to additional challenges that standard chemical structural
representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class
of compounds behave in a certain context in a biological system rather
than a single molecule with a fully-specified structure -- and for
similar reasons, chemical concepts such as \textit{groups} (parts of
molecules); secondly, the need to represent \textit{non-structural}
groupings of compounds of interest, since compounds may bind to the
same or similar targets or they may have similar biological functions
such as acting as an anti-neoplastic -- preventing the development of
tumors. With the many different biological databases needing to refer
to these different chemical entities in an organized and standardized
fashion to enable cross-database integration for a whole-systems
perspective, \textit{ontologies} and other semantic technologies are
being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within biological
systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, which contains around 27,100
entities in an interlinked semantic graph as of February 2012, and
is actively used for annotation in dozens of biological databases.
Ontologies are based on logical languages of varying levels of
expressivity, accompanied by sophisticated reasoning algorithms. They
can be used to construct various sorts of semantic similarity
(e.g. similarity in function, similarity in application) which
complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical structure
information with the wider logic-based ontology information, which
will allow combined automated reasoning and querying over the
different domains encoded in a growing number of interlinked
ontologies. 
Enabling automated inferencing in such heterogenous linked systems
brings us closer to achieving a systems level understanding of small
molecule activity \cite{Oprea:2007fk}.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} mentioned earlier, 
the goal of any modeling approach is to capture
and characterize correlations between structural features and
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using the
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input that can be
reliably predicted, usually delineated by similarity
to the training set.

@@

The first step in the predictive modeling of biological activities is
to generate \emph{molecular descriptors} (a.k.a. \emph{features}) that
are numerical representations of different structural features. For
example, labeled graphs and their associated
characterizations are easily accessible to computer scientists, yet
such features miss significant physicochemical features (
properties such as charge, flexibility and so on). At the same time,
it can be difficult to objectively quantify many chemical aspects of a
molecule, such that the resultant descriptors are suitable for
predictive modeling.  Hence, the choice of a chemical descriptor
should by no means be treated as a ``solved'' problem. For a more
detailed discussion the reader is referred to more comprehensive
textbooks \cite{todeschini2000,faulon2010}.

As noted above, molecular graphs can be transformed to numerical
vector representations ranging from counts of elements to eigenvalues
of the Laplacian matrix.  Alternatively, we can compare molecular
graphs directly, via \emph{kernel methods}, where a kernel on graphs
$G$ and $G'$ provides a measure of how similar $G$ is to $G'$, or a
kernel on a single graph compares measured similarities between the
graph's nodes. In these cases, rather than compute vector
representations, we directly operate on the graph representations.
Both methods have advantages and disadvantages. The vector approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem is
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels such as those based on multiple common substructures. 
These challenges can
make kernel based methods prohibitive on larger datasets.

Having settled on a numerical representation and a possible class of
model types, one must address the goal of the model. Are we looking
for pure predictive power, with no interest in any explanatory
features? Or are we more interested in decent predictive power, but
also some explanation as to why a molecule is predicted as toxic or
non-toxic (for example)? The former scenario is common in virtual
screening settings, where we require a high degree of accuracy and
fast predictions, but do not really care why one molecule is active
and another inactive. These types of models can be black boxes (e.g.,
neural networks) and algorithmic models \cite{Breiman:2001fk} (e.g., random
forests). The latter scenario is more common in exploratory and
optimization settings, where we hope that the output of the model will
guide us in chemical modifications to improve the property being
modeled.  In this case, we need to understand the effect of a certain
structural feature on the observed potency and expect that the model
will provide insight. These types of models are generally
distributional in nature (such as linear regression, Partial Least
Squares and Na\"{i'}ve Bayes) though some algorithmic approaches
 can provide insight.
